Modelling the Lymphatic Metastatic Progression Pathways of OPSCC from Multi-Institutional Datasets

### 2023-10-17 src/tex/ms.tex
[showyourwork] overleaf sync

% Author list
\author{Roman Ludwig}
\author{Jean-Marc Hoffmann}
\author{Bertrand Pouymayou}
\author{Lauence Bauwens}
\author{Vincent Grégoire}
\author{Roland Giger}
\author{Panagiotis Balermpas}
\author{Jan Unkelbach}

@@

To determine the optimal graph describing the data at hand best, we considered six models, each with one of the green candidate arcs of \cref{fig:extended_graph} added to the base graph. Every one of these models was evaluated as described below in \cref{subsec:complete_model:comparison}.

@@

@@

The dataset $\boldsymbol{\mathcal{D}}$ that we used for inference is comprised of the detailed reports on lymph node involvement patterns in \gls{opscc} patients treated at three different institutions in France and Switzerland: The \gls{clb} in Lyon (France), the \gls{isb}, and the \gls{usz} (both in Switzerland). We have previously published the patterns of nodal involvement for the \gls{usz} cohort (287 patients) \cite{ludwig_dataset_2022} and described its characteristics in detail \cite{ludwig_detailed_2022}. A large part of the \gls{clb} data (263 patients) underlies a publication on \gls{hpv} status in \gls{opscc} \cite{bauwens_prevalence_2021}. The data from \gls{clb} and \gls{isb} are described in details in a data set publication [lets cite it as under review and attach the submitted draft of the DiB].

In total, the dataset contains \variable{output/num_patients.txt} patients with newly diagnosed \gls{opscc}. It includes patients treated with definitive (chemo)radiotherapy, adjuvant (chemo)radiotherapy following neck dissection, or neck dissection alone. Pathologically assessed \gls{lnl} involvement was available for 393 surgically treated patients, while for the remainder the nodal involvement was assessed based on available diagnostic modalities (FDG-PET-CT, CT, MRI, FNA). If multiple modalities were used to diagnose a patient's lymph node involvement, the available modalities were combined into a consensus decision. When different modalities were conflicting, the conflicts were resolved by inferring the most likely state (healthy or metastatic) for each \gls{lnl} separately. To do so, we used literature values for the sensitivity and specificity of the diagnostic modalities \cite{de_bondt_detection_2007,kyzas_18f-fluorodeoxyglucose_2008}, which we also tabulated in \cref{table:sens_spec}. Practically, this means that, whenever pathology after neck dissection was available, the pathology result was taken as the consensus, overruling any other clinical diagnostic modality. If, for example, PET-CT and MRI ware available and conflicting, PET-CT was taken as the consensus, overruling MRI.

@@

@@

\Cref{table:evidence} reports the model evidence for all extended graphs that were investigated. The results is additionally visualized in Figure xxx. The winning graph adds arcs from level II to I, from level II to V, and from level III to level V to the base graph. Below, we further analyze why these connections improve the model's ability to describe the involvement patterns seen in the data.

>>>>

% Author list
\author{Roman Ludwig}
\author{Jean-Marc Hoffmann}
\author{Bertrand Pouymayou}
\author{Lauence Bauwens}
\author{Vincent Grégoire}
%\author{Roland Giger}
\author{... TBD ...}
\author{Panagiotis Balermpas}
\author{Jan Unkelbach}

@@

To determine the optimal graph describing the data at hand best, we considered six models, each with one of the green candidate arcs of \cref{fig:extended_graph} added to the base graph. Every one of these models was evaluated by computing its evidence via thermodynamic integration as described below in \cref{subsec:complete_model:comparison}. These evidences are shown in table XXX. The X most promising added edges were then considered in a combined model which was ultimately designated to be the "winning graph" which is in great detail compared to the base graph in the results (\cref{sec:results}). All considered graphs' performances are visualized in figure XXX.

@@

\rl{\textbf{NOTE:} This subsection will be updated based on the data that will be finally included.}

@@

%The dataset $\boldsymbol{\mathcal{D}}$ that we used for inference is comprised of the detailed reports on lymph node involvement patterns in \gls{opscc} patients treated at three different institutions in France and Switzerland: The \gls{clb} in Lyon (France), the \gls{isb}, and the \gls{usz} (both in Switzerland). We have previously published the patterns of nodal involvement for the \gls{usz} cohort (287 patients) \cite{ludwig_dataset_2022} and described its characteristics in detail \cite{ludwig_detailed_2022}. A large part of the \gls{clb} data (263 patients) underlies a publication on \gls{hpv} status in \gls{opscc} \cite{bauwens_prevalence_2021}. The data from \gls{clb} and \gls{isb} are described in details in a data set publication [lets cite it as under review and attach the submitted draft of the DiB].

The dataset $\boldsymbol{\mathcal{D}}$ that we used for inference is comprised of the detailed reports on lymph node involvement patterns in \gls{opscc} patients treated at two different institutions in France and Switzerland: The \gls{clb} in Lyon (France) and the \gls{usz} in Zürich (Switzerland). We have previously published the patterns of nodal involvement for the \gls{usz} cohort (287 patients) \cite{ludwig_dataset_2022} and described its characteristics in detail \cite{ludwig_detailed_2022}. The \gls{clb} data (263 patients) underlies a publication on \gls{hpv} status in \gls{opscc} \cite{bauwens_prevalence_2021}. \rl{\textbf{NOTE:} For the ISB and second CLB dataset, we can cite the data in brief paper as under review}.

In total, the dataset contains \variable{output/num_patients.txt} patients with newly diagnosed \gls{opscc}. It includes patients treated with definitive (chemo)radiotherapy, adjuvant (chemo)radiotherapy following neck dissection, or neck dissection alone. Pathologically assessed \gls{lnl} involvement was available for 263 surgically treated patients, while for the remainder the nodal involvement was assessed based on available diagnostic modalities (FDG-PET-CT, CT, MRI, FNA). If multiple modalities were used to diagnose a patient's lymph node involvement, the available modalities were combined into a consensus decision. When different modalities were conflicting, the conflicts were resolved by inferring the most likely state (healthy or metastatic) for each \gls{lnl} separately. To do so, we used literature values for the sensitivity and specificity of the diagnostic modalities \cite{de_bondt_detection_2007,kyzas_18f-fluorodeoxyglucose_2008}, which we also tabulated in \cref{table:sens_spec}. Practically, this means that, whenever pathology after neck dissection was available, the pathology result was taken as the consensus, overruling any other clinical diagnostic modality. If, for example, PET-CT and MRI ware available and conflicting, PET-CT was taken as the consensus, overruling MRI.

@@

\subsection{Graph Candidates}
\label{subsec:results:graph_candidates}

\rl{\textbf{NOTE:} This subsection is still unfinished an will be completed when the respective computations are ready. Qualitatively, we don't expect any major changes though.}

The model evidence of the six models that were based on graphs with one additional arc compared to the base model's is plotted in \cref{table:evidence}. A visual ranking is provided in figure XXX. Evidently, adding a connection from \gls{lnl} II to I is strongly/substantially supported given this dataset. Similarly, there is strong/substantial evidence for introducing an arc from \gls{lnl} II to V. All other investigated additions lead to improvements that are at best barely worth a mention or do not justify the additional complexity by yielding a higher accuracy at all.

Based on this table, we determined the candidate for the optimal graph to have both the new arc from level II to I, as well as arcs from \gls{lnl} II and \gls{lnl} III to level V. To ensure the evidence gain of these edges added individually carries over into one combined model, we performed another thermodynamic integration.

@@
